Decompositions of congruence subgroups of Chevalley groups

### 2015-10-09 paper.tex
Finished proofreading 5.2

It is classically known that over an arbitrary field $F$ the group $\SL_{n+1}(F)$, $n\geq 3$ is a product of at most $4$ subgroups of type $\SL(n, F)$.
On the other hand, one immediately obtains from \cref{thm:DennisVaserstein} that $\SL_{n+1}(R, I)$ factors into a product of a most $5$ subgroups of type $\SL_n(R, I)$ for any ideal $I$ satisfying $\sr(I)\leqslant n-1$.

>>>>

It is classically known that over an arbitrary field $F$ the group $\SL_{n+1}(F)$, $n\geq 3$ is a product of at most $4$ subgroups of type $\SL(n, F)$.
On the other hand, one immediately obtains from \cref{thm:DennisVaserstein} and surjective stability for linear $\K_1$ that $\SL_{n+1}(R, I)$ factors into a product
of a most $5$ subgroups of type $\SL_n(R, I)$ for any ideal $I$ satisfying $\sr(I)\leqslant n-1$.

### 2015-10-09 subsystem-factorizations.tex
"Varying v and w..." part
Finished proofreading 5.2

%The main result of \cite{NikProdDecomp} is the following
%\begin{thm*}
%Let $G$ be a classical (possibly twisted) Chevalley group of rank $n$ over a finite field. Then $G$ equals the product of at most $200$ conjugates of an $\SL_n$ subgroup.
%\end{thm*}
As indicated in the introduction, the Dennis---Vaserstein decomposition gives the following result.
\begin{lemma}
Assume $\sr(I)\leqslant n-1$. Then $\SL(n+1,R,I)$ is a product of at most $5$ subgroups isomorphic to $\SL(n,R,I)$.
\end{lemma}
\begin{proof}
By the Dennis---Vaserstein decomposition one can present $\SL(n+1,R,I)$ as a product
\[ \SL(n+1,R,I) =  \Par_1\cdot X_{n1}\cdot\Par_n=S_1\U_1\cdot X_{n1}\cdot\U_n S_n, \]
where $S_1$ and $S_n$ are two obvious embeddings of $\SL(n,R,I)$ in $\SL(n+1,R,I)$, avoiding respectively the first and the last row and column. Now $\U_1=(\U_1\cap S_n)\cdot X_{1n}$ and $U_n= X_{1n}\cdot(U_n\cap S_1)$, while $X_{1n}X_{n1}X_{1n}\in S_1^{w_{12}(1)}$.
\end{proof}
We will now elaborate on the case $\Phi=\rD_\ell$ to show that the Dennis---Vaserstein decomposition is suitable for handling other Chevalley groups, albeit with a stronger assumption on the base ring.

@@

The following lemma is an analogue of Proposition~1 of~\cite{NikProdDecomp}.
\begin{lemma}\label{lemma:nikolov-type-dl}
Let $\Phi=\rD_\ell$, $\ell\geq 4$.
There exist an element $y\in\E(\Phi,R)$ and an element $w\in\widetilde{W}(\Phi)$ such that $\U(\Sigma_\ell, I)\subset[\U(\Delta_\ell^-, I),y]\cdot{}^w\!\U(\Delta_\ell^+, I)$.
\end{lemma}
\begin{proof}
Since $\U(\Sigma_\ell, I)$ is abelian we can decompose it as $\U(\Sigma_\ell, I)=\U(\Sigma_\ell\setminus B, I) \cdot \U(B, I)$. 
Set $y=\prod_{\beta\in B}x_\beta(1)$. We will now show that $\U(\Sigma_\ell\setminus B, I)\subset[\U(\Delta_\ell^-, I),y]\cdot\U(B, I)$.

Notice that $\beta_1$ is the only root of $\Phi$ satisfying $m_2(\beta_1)=2$ therefore the commutator formula~\eqref{rel:CCF} implies
\[ \bigl[\U(\Delta_2^-, I),x_{\beta_1}(1)\bigr]=1. \]
Since the set $A=(\Sigma_2^-\cap\Delta_\ell)+(B\setminus\{\beta_1\})$ is disjoint from $\Phi$ 
(for $\alpha\in A$ one has $m_2(\alpha)=-1$ and $m_\ell(\alpha)=1$)
we also get from~~\eqref{rel:CCF} that
\[\Bigl[\U(\Sigma_2^-\cap\Delta_\ell, I),\prod_{i\neq1}x_{\beta_i}(1)\Bigr]=1. \]
Since $B\setminus\{\beta_1\}\subset\Sigma_\ell\cap\Delta_2$, the above identities imply
\[ \Bigl[ \U(\Sigma_2^-\cap\Delta_\ell, I)\cdot\U(\Delta_{2,\ell}^-, I), x_{\beta_1}(1)\cdot\prod_{i\neq1}x_{\beta_i}(1) \Bigr] \equiv \bigl[ \U(\Sigma_2^-\cap\Delta_\ell, I), x_{\beta_1}(1) \bigr] \bmod \U(\Sigma_\ell\cap\Delta_2, I). \]
Take an element $u\in\U(\Sigma_2^-\cap\Delta_\ell, I)$ and decompose it as $u=vw$, $v\in\U(\Sigma_1^-\cap\Sigma_2^-\cap\Delta_\ell, I)$, $w\in\U(\Sigma_2^-\cap\Delta_{1,\ell}, I)$.
Using the identity
\begin{equation}\label{eq:comm-ab-c}
[ab,c]={}^a[b,c]\cdot[a,c],
\end{equation}
we can rewrite
\[ [vw,x_{\beta_1}(1)] = {}^v[w,x_{\beta_1}(1)]\cdot[v,x_{\beta_1}(1)].  \]
Since $\U(\Sigma_1^-\cap\Sigma_2^-\cap\Delta_\ell, I)$ and $\U(\Sigma_2^-\cap\Delta_{1,\ell}, I)$ are abelian, it is easy to see that
\[ [v,x_{\beta_1}(1)]\in\U(\Sigma_2\cap\Sigma_\ell\cap\Delta_1, I), \quad [w,x_{\beta_1}(1)]\in\U((\Sigma_1\cap\Sigma_\ell)\setminus\{\beta_1\}, I). \]
%TODO: <add text from thesis>
Varying $v$ and $w$ one can obtain any element of $\U(\Sigma_2\cap\Sigma_\ell\cap\Delta_1)$ and $\U((\Sigma_1\cap\Sigma_\ell)\setminus\{\beta_1\})$ respectively. This is easy to see by using \eqref{eq:comm-ab-c} repeatedly and the fact that $\U(\Sigma_2^-\cap\Delta_\ell)$ commutes with $\U(\Sigma_\ell)$.
Now by the Levi decomposition
\[ {}^v\!\U((\Sigma_1\cap\Sigma_\ell)\setminus\{\beta_1\}) \equiv \U((\Sigma_1\cap\Sigma_\ell)\setminus\{\beta_1\}) \bmod \U(\Sigma_\ell\cap\Delta_2). \]
Thus we have shown that
\[ [\U(\Sigma_2^-)\cdot\U(\Delta_{2,\ell}),y] \equiv \U((\Sigma_{1,2}\cap\Sigma_\ell)\setminus\{\beta_1\}) \bmod \U(\Sigma_\ell\cap\Delta_2), \]
and this reduces the problem to the $\rD_{\ell-2}$-subsystem $\Delta_{1,2}$. So we can carry the induction on the rank of the subsystem, and construct for any given $a\in\U(\Sigma_\ell\setminus B)$ an element $b\in\U(\Delta_\ell^-)$ such that $a\in[b,y]\cdot\prod_{\beta\in B}X_\beta\subset[\U(\Delta_\ell^-),y]\cdot\U(B)$.
%TODO: </add text from thesis>

It remains to notice that the assertion of the lemma follows from~\cref{lemma:nikolov-weyl}.
%It remains to note that the roots $\pm\beta_i$ lie inside a maximal root subsystem of type $\rA_1+\ldots+\rA_1+\rD_3$.
%All such subsystems are conjugate under the action of the Weyl group, so there exist an element $w\in W(\Phi)$ such that
%$w(\beta_1)=\alpha_1$, $w(\beta_2)=\alpha_3$, $w(\beta_3)=\alpha_5$ and so on, so that all of 
%$w(\beta_i)$ lie in $\Delta_\ell^+$.
\end{proof}
\begin{proof}[Proof of \cref{thm:spin-sln-prod}]
By \cref{thm:DennisVaserstein} one has a decomposition
\[ \Spin(2\ell,R,I) = \Par_{\ell-1}\cdot\U_{\ell-1,\ell}^-\cdot\Par_{\ell} = \G(\Delta_{\ell-1})\cdot\U(\Sigma_{\ell-1})\cdot\U(\Sigma_{\ell-1,\ell}^-)\cdot\U(\Sigma_{\ell})\cdot\G(\Delta_\ell). \]
By \cref{lemma:nikolov-type-dl} there exist elements $y_1,y_2,y_3\in\G(\rD_\ell,R)$ and $w_1,w_2,w_3\in\widetilde{W}(\rD_\ell)$ such that
\begin{align*}
& \U(\Sigma_{\ell-1}) \subset \U(\Delta_{\ell-1}^-) \cdot {}^{y_1}\!\U(\Delta_{\ell-1}^-) \cdot {}^{w_1}\!\U(\Delta_{\ell-1}^+), \\
& \U(\Sigma_\ell) \subset {}^{w_2}\!\U(\Delta_\ell^+) \cdot {}^{y_2}\!\U(\Delta_\ell^-) \cdot \U(\Delta_\ell^-), \\
& \U(\Sigma_{\ell-1,\ell}^-) \subset \U(\Delta_\ell^+) \cdot {}^{y_3}\!\U(\Delta_\ell^+) \cdot {}^{w_3}\!\U(\Delta_\ell^-).
\end{align*}
Thus $\G(\rD_\ell,R,I)$ is a product of $9$ subgroups isomorphic to $\G(\rA_{\ell-1},R,I)\cong\SL(\ell,R,I)$.
\end{proof}

>>>>

%The main result of \cite{NikProdDecomp} is the following
%\begin{thm*}
%Let $G$ be a classical (possibly twisted) Chevalley group of rank $n$ over a finite field. Then $G$ equals the product of at most $200$ conjugates of an $\SL_n$ subgroup.
%\end{thm*}
%As indicated in the introduction, the Dennis---Vaserstein decomposition gives the following result.
%\begin{lemma}
%Assume $\sr(I)\leqslant n-1$. Then $\SL(n+1,R,I)$ is a product of at most $5$ subgroups isomorphic to $\SL(n,R,I)$.
%\end{lemma}
%\begin{proof}
%By the Dennis---Vaserstein decomposition one can present $\SL(n+1,R,I)$ as a product
%\[ \SL(n+1,R,I) =  \Par_1\cdot X_{n1}\cdot\Par_n=S_1\U_1\cdot X_{n1}\cdot\U_n S_n, \]
%where $S_1$ and $S_n$ are two obvious embeddings of $\SL(n,R,I)$ in $\SL(n+1,R,I)$, avoiding respectively the first and the last row and column. Now $\U_1=(\U_1\cap S_n)\cdot X_{1n}$ and $U_n= X_{1n}\cdot(U_n\cap S_1)$, while $X_{1n}X_{n1}X_{1n}\in S_1^{w_{12}(1)}$.
%\end{proof}
%We will now elaborate on the case $\Phi=\rD_\ell$ to show that the Dennis---Vaserstein decomposition is suitable for handling other Chevalley groups, albeit with a stronger assumption on the base ring.

@@

The following lemma is an analogue of Proposition~1 of~\cite{NikProdDecomp}.
\begin{lemma}\label{lemma:nikolov-type-dl}
Let $\Phi=\rD_\ell$, $\ell\geq 2$ and let $R$ be arbitrary ring.
There exist an element $y\in\E(\Phi,R)$ and an element $w\in\widetilde{W}(\Phi)$ such that $\U(\Sigma_\ell, I)\subset[\U(\Delta_\ell^-, I),y]\cdot{}^w\!\U(\Delta_\ell^+, I)$.
\end{lemma}
\begin{proof}
Since $\U(\Sigma_\ell, I)$ is abelian we can decompose it as $\U(\Sigma_\ell, I)=\U(\Sigma_\ell\setminus B, I) \cdot \U(B, I)$. 
Set $y=\prod_{\beta\in B}x_\beta(1)$. 
We will now show by induction on $\ell$ that 
\begin{equation}\label{eq:ind-stat} \U(\Sigma_\ell\setminus B, I)\subset[\U(\Delta_\ell^-, I),y]\cdot\U(B, I). \end{equation}
The induction base in the cases $\ell=2,3$ is trivial.

Notice that $\beta_1$ is the only root of $\Phi$ satisfying $m_2(\beta_1)=2$ therefore the commutator formula~\eqref{rel:CCF} implies
\[ \bigl[\U(\Delta_2^-, I),x_{\beta_1}(1)\bigr]=1. \]
Since the set $A=(\Sigma_2^-\cap\Delta_\ell)+(B\setminus\{\beta_1\})$ is disjoint from $\Phi$ 
(for $\alpha\in A$ one has $m_2(\alpha)=-1$ and $m_\ell(\alpha)=1$)
we also get from~~\eqref{rel:CCF} that
\[\Bigl[\U(\Sigma_2^-\cap\Delta_\ell, I),\prod_{i\neq1}x_{\beta_i}(1)\Bigr]=1. \]
Since $B\setminus\{\beta_1\}\subset\Sigma_\ell\cap\Delta_2$, the above identities imply
\[ \Bigl[ \U(\Sigma_2^-\cap\Delta_\ell, I)\cdot\U(\Delta_{2,\ell}^-, I), x_{\beta_1}(1)\cdot\prod_{i\neq1}x_{\beta_i}(1) \Bigr] \equiv \bigl[ \U(\Sigma_2^-\cap\Delta_\ell, I), x_{\beta_1}(1) \bigr] \bmod \U(\Sigma_\ell\cap\Delta_2, I). \]
Take an element $u\in\U(\Sigma_2^-\cap\Delta_\ell, I)$ and decompose it as $u=vw$, $v\in\U(\Sigma_1^-\cap\Sigma_2^-\cap\Delta_\ell, I)$, $w\in\U(\Sigma_2^-\cap\Delta_{1,\ell}, I)$.
Using the identity
\begin{equation}\label{eq:comm-ab-c}
[ab,c]={}^a[b,c]\cdot[a,c],
\end{equation}
we can rewrite
\[ [vw,x_{\beta_1}(1)] = {}^v[w,x_{\beta_1}(1)]\cdot[v,x_{\beta_1}(1)].  \]
Since $\U(\Sigma_1^-\cap\Sigma_2^-\cap\Delta_\ell, I)$ and $\U(\Sigma_2^-\cap\Delta_{1,\ell}, I)$ are abelian, it is easy to see that
\[ [v,x_{\beta_1}(1)]\in\U(\Sigma_2\cap\Sigma_\ell\cap\Delta_1, I), \quad [w,x_{\beta_1}(1)]\in\U((\Sigma_1\cap\Sigma_\ell)\setminus\{\beta_1\}, I). \]
Every element of $\U(\Sigma_2\cap\Sigma_\ell\cap\Delta_1, I)$ (resp. $\U(\Sigma_1\cap\Sigma_\ell\setminus\{\beta_1\}, I)$) can be obtained as above for a suitable choice of $v$ (resp. $w$).
Indeed, set $v=x_\gamma(\xi_\gamma)\cdot v'$, $\gamma=-\alpha_1-\alpha_2$, $v' \in \U(\Sigma_1^- \cap \Sigma_2^- \cap \Delta_\ell \setminus \{\gamma\})$.
Using relation \eqref{eq:comm-ab-c} and the fact that $\U(\Sigma_2^-\cap\Delta_\ell)$ commutes with $\U(\Sigma_\ell)$ we get that:
\begin{multline*} [v, x_{\beta_1}(1)] = [x_\gamma(\xi_\gamma)\cdot v', x_{\beta_1}(1)] = {}^{x_\gamma(\xi_\alpha)}[v', x_{\beta_1}(1)] \cdot [x_\gamma(\xi_\gamma), x_{\beta_1}(1)] = \\
 = [v', x_{\beta_1}(1)]\cdot x_{\beta_1-\alpha_1-\alpha_2}(\xi_\gamma) = \ldots = \prod_{\mathclap{\alpha\in\Sigma_1^- \cap \Sigma_2^- \cap \Delta_\ell}} x_{\beta_1+\alpha}(\xi_\alpha). \end{multline*}
It remains to note that $\Sigma_2 \cap \Sigma_\ell \cap \Delta_1 = \beta_1 + \Sigma_1^- \cap \Sigma_2^- \cap \Delta_\ell$. The same argument works for $[w, x_{\beta_1}(1)]$.
Direct calculation using the commutator formula shows that
\[ {}^v\!\U(\Sigma_1\cap\Sigma_\ell\setminus\{\beta_1\}, I) \equiv \U(\Sigma_1\cap\Sigma_\ell\setminus\{\beta_1\}, I) \bmod \U(\Sigma_\ell\cap\Delta_2, I). \]
Summing up the above arguments, we get that
\[ [\U(\Sigma_2^-, I)\cdot\U(\Delta_{2,\ell}^-, I),y] \equiv \U((\Sigma_{1,2}\cap\Sigma_\ell)\setminus\{\beta_1\}) \bmod \U(\Sigma_\ell\cap\Delta_2, I). \]
hence \eqref{eq:ind-stat} follows from the induction hypothesis (applied to $\Delta_{1,2} \cong \rD_{\ell-2}$).

Finally, we have found $a\in\U(\Sigma_\ell\setminus B, I)$ and $b\in\U(\Delta_\ell^-, I)$
such that $$a\in[b,y]\cdot\prod_{\beta\in B}X_\beta\subset[\U(\Delta_\ell^-, I),y]\cdot\U(B, I).$$
Now the assertion of the lemma follows from~\cref{lemma:nikolov-weyl}.
\end{proof}

\begin{proof}[Proof of \cref{thm:spin-sln-prod}]
Set $L = \E(\Delta_\ell, R, I) \leq \E(\rD_\ell, R, I)$ and denote by $\sigma$ the automorphism of $\G(\rD_\ell, R)$ induced by the sole diagram automorphism of the Dynkin diagram of $\rD_\ell$.
By \cref{thm:DennisVaserstein} one has
\begin{multline*}
\E(\rD_\ell,R,I) = \EP_{\ell}(R, I)\cdot\U(\Sigma_{\ell-1}^- \cap \Sigma_{\ell}^-, I)\cdot\EP_{\ell-1}(R, I) = \\
= L \cdot\U(\Sigma_{\ell}, I)\cdot\U(\Sigma_{\ell-1} \cap \Sigma_{\ell}^-, I)\cdot \left(L \cdot \U(\Sigma_{\ell}, I) \right)^\sigma.
\end{multline*}  
Now using \cref{lemma:nikolov-type-dl} one can find $y_1,y_2\in\G(\rD_\ell,R)$ and $w_1,w_2\in\widetilde{W}(\rD_\ell)$ such that
\begin{align*} & L \cdot \U(\Sigma_{\ell}, I) \subset L \cdot \U(\Delta_{\ell-1}^-, I) \cdot {}^{y_1}\!\U(\Delta_{\ell-1}^-, I) \cdot {}^{w_1}\!\U(\Delta_{\ell-1}^+, I), \\
& \U(\Sigma_{\ell-1,\ell}^-, I) \subset \U(\Delta_\ell^+, I) \cdot {}^{y_2}\!\U(\Delta_\ell^+, I) \cdot {}^{w_2}\!\U(\Delta_\ell^-, I). \end{align*} 
Thus $\E(\rD_\ell,R,I)$ is a product of at most $9$ conjugates of $L \cong \E(\rA_{\ell-1}, R, I)$. \end{proof}
